Corpus Experiments for Word Embeddings

### 2015-08-18 article/main.tex
generate three sample texts for the article

% compile using latex (xelatex seems to crop the figure strangley)
\documentclass{article} % For LaTeX2e
\usepackage[dvips]{graphicx}
\usepackage{nips14submit_e,times}
\usepackage{hyperref}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[outdir=./]{epstopdf}

@@

\newcommand{\p}{\text{P}}
\newcommand{\word}[1]{\texttt{#1}}
\urldef{\googlegroup}\url{https://groups.google.com/forum/#\!forum/word2vec-toolkit}
\urldef{\blogpost}\url{https://blog.lateral.io/2015/06/the-unknown-perils-of-mining-wikipedia/ }

@@

\section{Introduction}
Word embeddings are typically trained with samples from the word co-occurrence distributions.
Recall that the co-occurrence distribution of a word $w$ gives the probability $\p(w'|w)$ that a word $w'$ occurs nearby, given that $w$ occurred.
Samples of the co-occurrence distributions are obtained by scanning a short window over the text.
The non-central (or ``context'') words are treated as samples of the co-occurrence distribution the middle (or ``current'') word. 
In this way, the word vectors of the word embedding are determined by the co-occurrence distributions and frequencies of the words in the vocabulary.

@@

We illustrate our approach in case of the word2vec CBOW model.
However, these experiments (and other besides) could might equally well be performed for word embedding methods such as word2vec skip-gram, GloVe and SENNA.  \cite{}
In the case of word2vec, we show that word vector length depends directly and linearly with word frequency.
We show furthermore that word vector length depends directly and linearly with the level of noise in the co-occurrence distribution of the word.
In both cases, the coefficient of linearity depends upon the word.

In the word2vec similarity and word relationship tasks, normalised word vectors are used.
Word vector length is thus disregarded.
In the word2vec forum\footnote{\googlegroup}, it has been observed that normalisation improves performance on theses tasks since word vector length is related to word frequency.
It remained unclear, however, whether word frequency was \textit{directly} related to word vector lengthor rather via correlated factors such as word significance.
Our experimental approach demonstrates that a direct relationship indeed exists.
It is shown moreover that the normalisation of the word vectors also prevents word co-occurrence noise from affecting the performance of the word similarity and word relationship tasks.

\section{Model and corpus}
Our training data is built from the Wikipedia XML datadump from October 2013.
In order to remove from the training data the bulk of robot-generated pages on Wikipedia, only pages with at least 20 monthly page views are retained\footnote{For further justification of this rationale, and to obtain the dataset, see \blogpost}.
Stubs and disambiguation pages are also removed, leaving 463 thousand pages with a total of 482 million words.

@@

For simplicity, only the CBOW method with a single set of hyperparameters is considered.
Specifically, a $100$-dimensional CBOW model is trained using negative sampling with 5 negative samples, a window size of $10$, a minimum word occurrence of $200$, and $10$ passes through the corpus.
Subsampling was not used so that influence of word frequency could be more clearly discerned.
Identical experimental results were obtained using hierarchical softmax, but these are omitted for succinctness.

@@

\section{Experimental design and corpus modifications}\label{experimental_design}
Two controlled experiments are performed:
\begin{itemize}
	\item \emph{Word frequency variation} compare the lengths of the word vectors for words with varying frequencies, but with the same co-occurrence distribution;
	\item \emph{Co-occurrence noise variation} compare the lengths of the word vectors for words with of the same frequency, but with varying levels of noise in their co-occurrence distributions.
\end{itemize}

To perform these experiments, tokens are introduced into the corpus described in subsections \ref{WFVE} and \ref{CNVE}.
These tokens are introduced via a replacement procedure.
To illustrate this procedure, consider a word, e.g. \word{cat}.
For each occurrence of this word, a sample $i$, $1 \leqslant i \leqslant n$ is drawn from a truncated geometric distribution, and that occurrence of \word{cat} is replaced with \word{CAT\_i}.
Thus the word \word{cat} is replaced throughout the corpus by a family of tokens with varying frequencies but approximately the same co-occurrence distribution as \word{cat}.

The geometric distribution is truncated in order to limit the number of tokens introduced into the corpus.
For any ratio $0 < \lambda < 1$ and maximum value $n > 0$, the truncated geometric distribution is given by
$$ P_{\lambda, n} (i) = \frac{\lambda^{i-1} \cdot (1-\lambda)}{(1 - \lambda^n)}, \qquad 1 \leqslant i \leqslant n.$$ 
This is the distribution over $i = 1, 2, \dots n$ for which the probabilities decay exponentially base $\lambda$.
Of course, other distributions might equally have been chosen for the experiments.

\subsection{Word frequency variation experiment}\label{WFVE}
Firstly, a small set of words from the unmodified corpus is chosen uniformly at random from the vocabulary.
In order that the introduced tokens do not have too low a frequency, only words which occur at least 10 thousand times are chosen.
To this set of words we add the word \word{the}, in order to include a high-frequency stopword.
The replacement procedure is performed for each of these words, using a geometric rate of decay of $\lambda = 0.5$, and maximum value $n=20$.

Secondly, in order to have a meaningless word for comparison, an entirely new word \word{VOID} is interspersed uniformly at random throughout the corpus so that its global frequency is $0.005$.
The replacement procedure is then performed for this word with the same values for $\lambda$ and $n$.

Figure \ref{fig:word-frequency-experiment-text} shows the effect of these modifications on sample text, in the case where only the word \word{cat} is chosen.

\subsection{Co-occurrence noise variation experiment}\label{CNVE}
Here, and throughout, the noise distribution is taken to be the global word frequency distribution.
Thus noise can be added to the cooccurrence distribution of a word by inspersing occurrences of that word uniformly at randomly throughout the corpus.
A small set of words is chosen from the unmodified corpus in the same manner as subsection \ref{WFVE}.
In order that the two experiments can be carried out simulataneously, the two sets of words are chosen to be disjoint.

For each of the chosen words, the replacement procedure is performed using a geometric rate of decay of $\lambda = 5/6$, and truncating at $n=8$.
Subsequently, for every replacement token (e.g. \word{CAT\_i}) random occurrences of this token are interspersed uniformly at random throughout the corpus, such that the frequency of the replacement token is restored to that of the original token \word{cat}.
For example, if the original token \word{cat} occurred $1000$ times, then after the replacement procedure, \word{CAT\_2} occurs approximately $694$ times, so a further approximately $306 (=1000 - 694)$ random occurrences of \word{CAT\_2} are interspersed throughout the corpus.
Thus token \word{cat} is removed from the corpus, and a family of tokens \word{CAT\_i}, $1 \leqslant i \leqslant 8$ are introduced.
These tokens have the same frequency as \word{cat}, but their cooccurrence distributions, while based on that of \word{cat}, have an increasing amount of noise.
To be precise, the proportion of the occurrences of \word{CAT\_i} that arose by replacing \word{cat} is $P_{5/6}(i)$, and the proportion arising from its random intersperal throughout the corpus $(1 - P_{5/6}(i))$.

@@

\begin{figure*}
	%\input{../outputs/word-frequency-experiment-counts.tex}
\begin{center}
\begin{tabular}{c | c}
word & \# occurrences \\
\hline
\word{developments} & 14347 \\
\word{specific} & 77919 \\
\word{across} & 101617 \\
\word{path} & 24898 \\
\word{file} & 31184 \\
\word{from} & 2611891 \\
\word{admiral} & 13122 \\
\word{eyes} & 27621 \\
\end{tabular}
\end{center}
\label{fig:word-frequency-counts}
\caption{Occurrence counts for words chosen for the word frequency experiment. }
\end{figure*}

\begin{figure*}
	%\input{../outputs/word-frequency-experiment-counts.tex}
\begin{center}
\begin{tabular}{c | c}
word & \# occurrences \\
\hline
\word{sun} & 49723 \\
\word{jets} & 10520 \\
\word{holding} & 31744 \\
\word{incorporated} & 24964 \\
\word{initiative} & 15635 \\
\word{jury} & 14575 \\
\word{afternoon} & 10537 \\
\word{plants} & 42808 \\
\end{tabular}
\end{center}
\label{fig:cooccurrence-noise-counts}
\caption{Occurrence counts for words chosen for the cooccurrence noise experiment. }
\end{figure*}

% text after modifications for word frequency variation experiment for the word cat
\begin{figure}[t]
\texttt{
a pedigreed \textcolor{red}{CAT\_2} is one whose ancestry is recorded by a \textcolor{red}{CAT\_5} fancier organization
a purebred \textcolor{red}{CAT\_7} is one whose ancestry contains only individuals of the same \textcolor{red}{VOID\_4} breed
the semiferal \textcolor{red}{CAT\_1} a mostly outdoor \textcolor{red}{CAT\_1} is not owned by \textcolor{red}{VOID\_7} any one individual
the domestic \textcolor{red}{CAT\_1} was first classified as felis catus
the domesticated \textcolor{red}{CAT\_1} and its closest wild ancestor are both diploid organisms that possess  chromosomes and roughly  genes
the \textcolor{red}{CAT\_5} skull is unusual among mammals in having very large eye sockets and a powerful and specialized jaw
when it overpowers its prey a \textcolor{red}{CAT\_3} delivers a lethal neck bite with its two long canine teeth
another unusual feature is that the \textcolor{red}{CAT\_1} cannot produce taurine
this is partly the result of \textcolor{red}{CAT\_1} eyes having a tapetum lucidum which reflects any light that passes through the retina back into the eye
although wildcats are solitary the social behavior of domestic cats is much more variable and ranges from widely dispersed individuals to feral \textcolor{red}{CAT\_1} colonies that form around a food source based on groups of cooperating females
within such groups one \textcolor{red}{CAT\_1} is usually dominant over the others
each \textcolor{red}{CAT\_1} in a colony holds a distinct territory with sexually active males having the largest territories which are about  times larger than those of female cats and may overlap with several females territories
these territories are marked by urine spraying by rubbing objects at head height with secretions from facial glands and by defecation
between these territories are neutral areas where cats watch and greet one another without territorial conflicts
outside these neutral areas territory holders usually chase away stranger \textcolor{red}{VOID\_2} cats at first by staring hissing and growling and if that does not work by short but noisy and violent attacks
despite some cats cohabiting in colonies they do not have a social survival strategy or a pack mentality and always hunt alone
the domestic dog canis lupus familiaris or canis familiaris is a domesticated canid which has been selectively bred for millennia for various behaviors sensory capabilities and physical attributes
although initially thought to have originated as a manmade variant of an extant canid species variously supposed as being the dhole golden jackal or gray wolf extensive genetic studies undertaken during the s indicate that dogs diverged from a now-extinct canid in eurasia  years ago being the oldest domesticated animals their long association with people has allowed dogs to be uniquely attuned to human behavior as well as thrive on a starch-rich diet which would be inadequate for other canid species
dogs perform many roles for people such as hunting herding pulling loads protection assisting police and military companionship and more recently aiding handicapped individuals in  carl linnaeus listed among the types of quadrupeds familiar to him the latin word for \textcolor{red}{VOID\_3} dog canis among the species within this genus linnaeus listed the red fox as canis vulpes wolves canis lupus and the domestic dog canis canis in later editions linnaeus dropped canis canis and greatly expanded his list of the canis genus of quadrupeds and by  included alongside the foxes wolves and jackals and many more terms that are now listed as synonyms for domestic dog including aegyptius hairless dog aquaticus water dog and mustelinus literally badger dog among these were two that later experts have been widely used for domestic dogs as a species: canis domesticus and most predominantly canis familiaris the common or familiar dog
}
\caption{Text modified for the word frequency experiment, where the word \word{cat} was chosen, $\lambda=0.5$ and $n=20$.}
\label{fig:word-frequency-experiment-text}
\end{figure}

% original text after modifications for the noise experiment with n=3 lambda 5/ 6 for cat
\begin{figure}[t]
	\texttt{
a pedigreed \textcolor{red}{CAT\_1} is \textcolor{red}{CAT\_1} one whose ancestry is recorded by a \textcolor{red}{CAT\_1} fancier organization
a purebred \textcolor{red}{CAT\_3} is one whose ancestry contains only individuals of the same breed
the semiferal \textcolor{red}{CAT\_1} a mostly outdoor \textcolor{red}{CAT\_3} is not owned by any one individual
\textcolor{red}{CAT\_1} the domestic \textcolor{red}{CAT\_3} was first classified as felis catus
the domesticated \textcolor{red}{CAT\_1} and its closest wild ancestor are both diploid organisms \textcolor{red}{CAT\_3} that possess  chromosomes and roughly  genes
the \textcolor{red}{CAT\_2} skull is unusual among mammals in having very large eye sockets and a powerful and specialized jaw
when it overpowers its \textcolor{red}{CAT\_2} prey a \textcolor{red}{CAT\_1} delivers a lethal neck bite with \textcolor{red}{CAT\_3} its two \textcolor{red}{CAT\_2} long canine teeth
another unusual feature \textcolor{red}{CAT\_1} is that the \textcolor{red}{CAT\_2} cannot produce taurine
this is partly the result of \textcolor{red}{CAT\_3} eyes having a tapetum lucidum which reflects any light that passes through the retina back into the eye
although wildcats are solitary the social behavior of domestic cats is much more variable and \textcolor{red}{CAT\_1} ranges from widely dispersed individuals to feral \textcolor{red}{CAT\_2} colonies that form around a food source based on groups of cooperating females
within such groups one \textcolor{red}{CAT\_2} is usually dominant over the others
each \textcolor{red}{CAT\_1} in \textcolor{red}{CAT\_3} a colony holds a distinct \textcolor{red}{CAT\_3} territory with sexually active males having the largest territories which are about  times larger than those of female cats and may overlap with several females territories
these territories \textcolor{red}{CAT\_3} are marked by urine spraying by rubbing objects \textcolor{red}{CAT\_1} at head height with secretions from facial glands and by defecation
between these territories are neutral areas where cats \textcolor{red}{CAT\_1} watch and greet \textcolor{red}{CAT\_2} one another without territorial conflicts
\textcolor{red}{CAT\_1} outside these neutral areas territory holders usually chase away stranger cats at first by staring hissing and growling and if that does not work by short but \textcolor{red}{CAT\_2} noisy and violent attacks
despite some cats cohabiting in colonies they do \textcolor{red}{CAT\_3} not have a social survival strategy \textcolor{red}{CAT\_2} or a pack mentality and always hunt alone
the domestic dog canis lupus familiaris or canis familiaris is a domesticated canid which has been selectively bred for millennia for various behaviors sensory capabilities and physical attributes
although initially thought to have originated as a manmade variant of an extant canid species variously supposed as being the \textcolor{red}{CAT\_1} dhole golden jackal or gray wolf \textcolor{red}{CAT\_3} extensive genetic studies undertaken during the s indicate that dogs \textcolor{red}{CAT\_1} diverged from a now-extinct canid in eurasia  years ago being the oldest domesticated animals their long association with people has \textcolor{red}{CAT\_3} allowed dogs to be \textcolor{red}{CAT\_1} uniquely attuned to human behavior as well as thrive on \textcolor{red}{CAT\_2} a starch-rich diet which would be inadequate \textcolor{red}{CAT\_2} for other \textcolor{red}{CAT\_1} canid species
dogs perform many roles for people such \textcolor{red}{CAT\_3} as hunting herding pulling loads protection assisting \textcolor{red}{CAT\_2} police and military companionship and more recently aiding handicapped \textcolor{red}{CAT\_2} individuals in  carl linnaeus listed among the types of quadrupeds familiar to him the latin word for \textcolor{red}{CAT\_1} dog canis among \textcolor{red}{CAT\_1} the species within this genus linnaeus listed the red fox as canis vulpes wolves canis lupus and the domestic dog canis canis in later editions linnaeus dropped canis canis and greatly expanded his list of the canis genus of quadrupeds and by  included alongside the foxes wolves and jackals and \textcolor{red}{CAT\_2} many more terms that are now listed as synonyms for domestic dog including aegyptius \textcolor{red}{CAT\_1} hairless dog aquaticus water dog and mustelinus literally badger dog among these were two that later experts have been widely used for domestic dogs as a species: canis domesticus and most predominantly \textcolor{red}{CAT\_2} canis familiaris the common or familiar dog
}
\caption{Text modified for the co-occurrence noise experiment, where the word \word{cat} was chosen, $\lambda = 5/6$ and $n=3$.}
\label{fig:cooccurrence-noise-experiment-text}
\end{figure}

\section{Results}

\subsection{Word frequency experiment}
Figure \ref{fig:frequency-norm-graph} illustrates the complex global relationship between vector length and word frequency.
Figure \ref{fig:word-frequency-experiment-graph}, on the other hand, shows this relationship for individual words.
Each line corresponds to a single word, and the points on each line indicate the frequency and vector length for the tokens derived from the corresponding word.
For example, the five points on the line corresponding to \word{admiral} are labelled, from right to left, by the tokens \word{ADMIRAL\_1}, \word{ADMIRAL\_2}, \dots, \word{ADMIRAL\_5}.
The number of points on the line is determined by the frequency of the original word.
For example, the frequency of the \word{admiral} can be halved at most $5$ times and remain about the minimum frequency cut-off of word2vec.

\begin{figure*}
\includegraphics[scale=0.5]{occurrence-histogram}
\caption{
	Word frequency histogram.
	Here and elsewhere, frequency bands are delimited by powers of $2$.
}
\label{fig:frequency-histogram}
\end{figure*}

\begin{figure*}
\includegraphics[scale=0.3]{frequency-norm-scatterplot}
\caption{ The global relationship between word frequency and vector length.  }
\label{fig:frequency-norm-graph}
\end{figure*}

\begin{figure*}
\includegraphics[scale=0.5]{word-frequency-experiment-heatmap}
\caption{
Heatmap of the dot product of the normalised vectors associated to the tokens introduced into the corpus in the co-occurrence noise experiment (four such words chosen at random).
The largely red blocks demonstrate that the direction of the word vector does not change when noise is added to the co-occurrence distribution.
}

\label{fig:cooccurrence-noise-heatmap}
\end{figure*}
\begin{sidewaysfigure*}
	\centering{\includegraphics[scale=0.6]{word-frequency-experiment-graph}}
\caption{
	Vector length vs. frequency for the words chosen for the word frequency experiment.
	For each word, tokens of varying frequency but with the co-occurrence distribution of that word were introduced into the corpus, as described in section \ref{WFVE}.
}
\label{fig:word-frequency-experiment-graph}
\end{sidewaysfigure*}

\subsection{Co-occurrence noise experiment}
Consider the effect on a word vector when increasing levels of noise added to the co-occurrence distribution of the word.
This is affected by randomly dispersing an increasing proporition $0 \leqslant p \leqslant 1$ of the pre-existing occurrences of that word throughout the corpus.
The word vector depends on $p$, and thus $p$ parameterises a path in the feature space.
This path begins (with $p=0$) at the original word vector, and ends (with $p=1$) at the vector associated with the meaningless vector \word{VOID}.

@@

\begin{sidewaysfigure*}
\includegraphics[scale=0.6]{cooccurrence-noise-graph}
\caption{
	Vector length vs. proportion of co-occurrence noise for words chosen for the co-occurrence noise experiment.
	For each word, tokens of equal frequency but with increasing proporitions of co-occurrence noise were introduced into the corpus, as described in subsection \ref{CNVE}.
}
\label{fig:cooccurrence-noise-graph}
\end{sidewaysfigure*}

\begin{figure*}[t!]
\includegraphics[scale=0.5]{cooccurrence-noise-heatmap}
\caption{
Heatmap of the dot product of the normalised vectors associated to the tokens introduced into the corpus in the co-occurrence noise experiment (four such words chosen at random).
The largely red blocks demonstrate that the direction of the word vector does not change when noise is added to the co-occurrence distribution.
}
\label{fig:cooccurrence-noise-heatmap}
\end{figure*}

@@

\footnotesize
\bibliographystyle{unsrt}
\bibliography{main}

\end{document}

>>>>

% compile using latex (xelatex seems to crop the figure strangley)
\documentclass{article} % For LaTeX2e
\usepackage[dvips]{graphicx}
\usepackage{nips14submit_e,times}
\usepackage{hyperref}
%\usepackage{wrapfig}
\usepackage{natbib}
\usepackage{rotating}
\usepackage[outdir=./]{epstopdf}

@@

\newcommand{\p}{\text{P}}
\newcommand{\word}[1]{\texttt{#1}}
\urldef{\googlegroup}\url{https://groups.google.com/forum/#\!forum/word2vec-toolkit}
\urldef{\blogpost}\url{https://blog.lateral.io/2015/06/the-unknown-perils-of-mining-wikipedia/}

@@

\begin{section}{Introduction}
Word embeddings are typically trained with samples from the word co-occurrence distributions.
Recall that the co-occurrence distribution of a word $w$ gives the probability $\p(w'|w)$ that a word $w'$ occurs nearby, given that $w$ occurred.
Samples of the co-occurrence distributions are obtained by scanning a short window over the text.
The non-central (or ``context'') words are treated as samples of the co-occurrence distribution the middle (or ``current'') word. 
In this way, the word vectors of the word embedding are determined by the co-occurrence distributions and frequencies of the words in the vocabulary.

@@

We illustrate our approach in case of the word2vec CBOW model.
However, these experiments (and other besides) could might equally well be performed for word embedding methods such as word2vec skip-gram \cite{DistRepns,EfficientEstimation}, GloVe \cite{pennington2014glove} and SENNA \cite{collobert-2011}.
In the case of word2vec, we show that word vector length depends directly and linearly with word frequency.
We show furthermore that word vector length depends directly and linearly with the level of noise in the co-occurrence distribution of the word.
In both cases, the coefficient of linearity depends upon the word.

In the word2vec similarity and word relationship tasks, normalised word vectors are used.
Word vector length is thus disregarded.
In the word2vec forum\footnote{\googlegroup}, it has been observed that normalisation improves performance on theses tasks since word vector length is related to word frequency.
Figure \ref{fig:frequency-norm-graph} illustrates the global relationship between vector length and word frequency.
It remained unclear, however, whether word frequency was \textit{directly} related to word vector length or rather via correlated factors such as word significance.
Our experimental approach demonstrates that a direct relationship indeed exists.
It is shown moreover that the normalisation of the word vectors also prevents word co-occurrence noise from affecting the performance of the word similarity and word relationship tasks.

\begin{subsection}{Previous work}
Our experimental finding, in the context of word2vec, that word vector length decreases with co-occurrence noise is reminiscent of \cite{vecchi-baroni-zamparelli2011}, where a relationship between vector length and the ``semantic deviance'' of an adjective-noun composition was studied observationally.
In \cite{schakel-wilson}, the interpretation of word vector length is an indication of word significance is explored in the context of the arXiv high-energy physics corpus.
\end{subsection}

\begin{figure}
	\includegraphics[scale=0.3]{frequency-norm-scatterplot}
	\caption{ The global relationship between word frequency and vector length.  }
	\label{fig:frequency-norm-graph}
\end{figure}
\end{section}

\begin{section}{Corpus and model}
Our training data is built from the Wikipedia XML datadump from October 2013.
In order to remove from the training data the bulk of robot-generated pages on Wikipedia, only pages with at least 20 monthly page views are retained\footnote{For further justification of this rationale, and to obtain the dataset, see \blogpost}.
Stubs and disambiguation pages are also removed, leaving 463 thousand pages with a total of 482 million words.

@@

For simplicity, only the CBOW method with a single set of hyperparameters is considered.
Specifically, a $100$-dimensional CBOW model is trained using negative sampling with 5 negative samples, a window size of $10$, a minimum word occurrence of $200$, and $10$ passes through the corpus.
Subsampling was not used so that influence of word frequency could be more clearly discerned.
Identical experimental results were obtained using hierarchical softmax, but these are omitted for succinctness.
The relatively high minimum word cut-off is chosen to ensure that word vectors receive a sufficient number of gradient updates to be meaningful.

@@

\begin{figure}\label{fig:frequency-histogram}
	\includegraphics[scale=0.5]{occurrence-histogram}
	\caption{
	Word frequency histogram.
	Here and elsewhere, frequency bands are delimited by powers of $2$.
	}
\end{figure}

\begin{subsection}{Replacement procedure}\label{replacement-procedure}
In the experiments, tokens are introduced into corpus via a replacement procedure.
To illustrate this procedure, consider a word, e.g. \word{cat}.
For each occurrence of this word, a sample $i$, $1 \leqslant i \leqslant n$ is drawn from a truncated geometric distribution, and that occurrence of \word{cat} is replaced with \word{CAT\_i}.
Thus the word \word{cat} is replaced throughout the corpus by a family of tokens with varying frequencies but approximately the same co-occurrence distribution as \word{cat}.

The geometric distribution is truncated in order to limit the number of tokens introduced into the corpus.
For any ratio $0 < \lambda < 1$ and maximum value $n > 0$, the truncated geometric distribution is given by
$$ P_{\lambda, n} (i) = \frac{\lambda^{i-1} \cdot (1-\lambda)}{(1 - \lambda^n)}, \qquad 1 \leqslant i \leqslant n.$$ 
This is the distribution over $i = 1, 2, \dots n$ for which the probabilities decay exponentially base $\lambda$.
Of course, other distributions might equally have been chosen for the experiments.
\end{subsection}
\end{section}

\begin{section}{Word frequency variation experiment}\label{WFVE}
In this experiment, families of tokens are introduced into the corpus.
The tokens in each family vary in frequency but share a common co-occurrence distribution.
After model training, comparison of the resulting word vectors for these tokens reveals the independent effect of frequency variation on the word embedding.

\begin{subsection}{Tokens derived from existing words}
A small set of words from the unmodified corpus is chosen uniformly at random from the vocabulary.
In order that the introduced tokens do not have too low a frequency, only words which occur at least 10 thousand times are chosen.
To this set of words we add the word \word{the}, in order to include a high-frequency stopword.
The replacement procedure of subsection \ref{replacement-procedure} is then performed for each of these words, using a geometric rate of decay of $\lambda = 0.5$, and maximum value $n=20$.
This value of $\lambda$ is one of a range of values that ensure that, for each word, multiple tokens will be introduced with a number of occurrences sufficient to survive the chosen word2vec minimum occurrence cut-off of 200.  
A maximum value $n=20$ suffices for this choice of $\lambda$, since $2^{20 + \log_2{200}}$ exceeds the maximum occurrence of any token in the corpus. 
Figure \ref{fig:word-frequency-experiment-text-cat} illustrates the effect of these modifications on a sample text, with a family of tokens \word{CAT\_i}, derived from the word \word{cat}.

Figure \ref{fig:word-frequency-counts} gives the number of occurrences of the words chosen for this experiment.

\begin{figure}\label{fig:word-frequency-counts}
	\input{word-frequency-experiment-counts.tex}
	\caption{Occurrence counts for words chosen for the word frequency experiment. }
\end{figure}

\begin{figure}
	\input{word-frequency-experiment-text-cat.tex}
	\caption{Text modified for the word frequency experiment, where the word \word{cat} was chosen, $\lambda=0.5$ and $n=20$.}
	\label{fig:word-frequency-experiment-text-cat}
\end{figure}

\end{subsection}

\begin{subsection}{Tokens derived from an introduced, meaningless word}
	A high-frequency, purely meaningless word is included in the experiment for comparison.
	We choose to introduce a new, entirely meaningless token \word{VOID} into the corpus, rather than choose an existing word whose meaninglessness is only supposed.
	To achieve this, the token is interspersed uniformly at random throughout the corpus so that its global frequency is $0.005$.
	The co-occurrence distribution of this introduced token thus coincides with the global frequency distribution.
	The replacement procedure is then performed for \word{VOID}, using the same values for $\lambda$ and $n$ as above.
Figure \ref{fig:word-frequency-experiment-text-void} illustrates the effect of this modifications on a sample text.

\begin{figure}\label{fig:word-frequency-experiment-text-void}
	\input{word-frequency-experiment-text-void.tex}
	\caption{Text modified for the word frequency experiment, where the
	word \word{cat} was chosen, $\lambda=0.5$ and $n=20$.}
\end{figure}
\end{subsection}

\subsection{Experimental results}
Word frequency has no effect on word vector direction.
Figure \ref{word-frequency-experiment-heatmap} shows the cosine similarity of the pairs of word vectors for the tokens introduced into the corpus.
The cosine similarity measures the extent to which two vectors have the same direction, taking a maximum value of $1$ and a minimum value of $0$.
Notice that the word vectors associated to tokens derived from meaningful tokens have the same direction.
The same is true of the word vectors associated to the \word{VOID\_i} for $i$ \textcolor{red}{below $7$ or $8$}, but that the word vectors \word{VOID\_i} gradually diverge directionally for higher values of $i$.
This demonstrates the difficulty of learning the co-occurrence distribution of $\word{VOID}$, the global frequency distribution.
Because of its noisiness, a higher number of samples (occurrences) is required to learn this co-occurrence distribution.

It remains, therefore, to consider the effect of frequency variation on word vector length.
Figure \ref{fig:frequency-norm-graph} illustrates the global relationship between vector length and word frequency.
Figure \ref{fig:word-frequency-experiment-graph}, on the other hand, shows this relationship for individual words.
Each line corresponds to a single word, and the points on each line indicate the frequency and vector length for the tokens derived from the corresponding word.
For example, the five points on the line corresponding to \word{admiral} are labelled, from right to left, by the tokens \word{ADMIRAL\_1}, \word{ADMIRAL\_2}, \dots, \word{ADMIRAL\_5}.
The number of points on the line is determined by the frequency of the original word.
For example, the frequency of the \word{admiral} can be halved at most $5$ times and remain about the minimum frequency cut-off.

\begin{figure}
	\includegraphics[scale=0.5]{word-frequency-experiment-heatmap}
	\caption{
	Heatmap of the dot product of the normalised vectors associated to the
	tokens introduced into the corpus in the co-occurrence noise experiment
	(four such words chosen at random).  The largely red blocks demonstrate
	that the direction of the word vector does not change when noise is
	added to the co-occurrence distribution.
	}
	\label{fig:cooccurrence-noise-heatmap}
\end{figure}

\begin{sidewaysfigure*}\label{fig:word-frequency-experiment-graph}
	\centering{\includegraphics[scale=0.6]{word-frequency-experiment-graph}}
	\caption{
	Vector length vs. frequency for the words chosen for the word frequency
	experiment.  For each word, tokens of varying frequency but with the
	co-occurrence distribution of that word were introduced into the
	corpus, as described in section \ref{WFVE}.
	}
\end{sidewaysfigure*}

\end{section}

\begin{section}{Co-occurrence noise variation experiment}\label{CNVE}
Here, and throughout, the noise distribution is taken to be the global word frequency distribution.
Thus noise can be added to the cooccurrence distribution of a word by inspersing occurrences of that word uniformly at randomly throughout the corpus.
A small set of words is chosen from the unmodified corpus in the same manner as section \ref{WFVE}.

For each of the chosen words, the replacement procedure of subsection \ref{replacement-procedure} is performed using a geometric rate of decay of $\lambda = 5/6$, and truncating at $n=8$.
Subsequently, for every replacement token (e.g. \word{CAT\_i}) random occurrences of this token are interspersed uniformly at random throughout the corpus, such that the frequency of the replacement token is restored to that of the original token \word{cat}.
For example, if the original token \word{cat} occurred $1000$ times, then after the replacement procedure, \word{CAT\_2} occurs approximately $694$ times, so a further approximately $306 (=1000 - 694)$ random occurrences of \word{CAT\_2} are interspersed throughout the corpus.
Thus token \word{cat} is removed from the corpus, and a family of tokens \word{CAT\_i}, $1 \leqslant i \leqslant 8$ are introduced.
These tokens have the same frequency as \word{cat}, but their cooccurrence distributions, while based on that of \word{cat}, have an increasing amount of noise.
To be precise, the proportion of the occurrences of \word{CAT\_i} that arose by replacing \word{cat} is $P_{5/6}(i)$, and the proportion arising from its random intersperal throughout the corpus $(1 - P_{5/6}(i))$.

@@

\begin{figure}
	\input{noise-cooccurrence-experiment-counts.tex}
	\label{fig:cooccurrence-noise-counts}
	\caption{Occurrence counts for words chosen for the cooccurrence noise experiment. }
\end{figure}

% original text after modifications for the noise experiment with n=3 lambda 5/ 6 for cat
\begin{figure}
	\input{cooccurrence-noise-experiment.tex}
	\caption{Text modified for the co-occurrence noise experiment, where the word \word{cat} was chosen, $\lambda = 5/6$ and $n=3$.}
\label{fig:cooccurrence-noise-experiment-text}
\end{figure}

\begin{subsection}{Results}
Consider the effect on a word vector when increasing levels of noise added to the co-occurrence distribution of the word.
This is affected by randomly dispersing an increasing proporition $0 \leqslant p \leqslant 1$ of the pre-existing occurrences of that word throughout the corpus.
The word vector depends on $p$, and thus $p$ parameterises a path in the feature space.
This path begins (with $p=0$) at the original word vector, and ends (with $p=1$) at the vector associated with the meaningless vector \word{VOID}.

@@

\begin{figure}
	\includegraphics[scale=0.5]{cooccurrence-noise-heatmap}
	\caption{
	Heatmap of the dot product of the normalised vectors associated to the
	tokens introduced into the corpus in the co-occurrence noise experiment
	(four such words chosen at random).  The largely red blocks demonstrate
	that the direction of the word vector does not change when noise is
	added to the co-occurrence distribution.
	}
	\label{fig:cooccurrence-noise-heatmap}
\end{figure}

\begin{sidewaysfigure*}\label{fig:cooccurrence-noise-graph}
	\includegraphics[scale=0.6]{cooccurrence-noise-graph}
	\caption{
	Vector length vs. proportion of co-occurrence noise for words
	chosen for the co-occurrence noise experiment.  For each word,
	tokens of equal frequency but with increasing proporitions of
	co-occurrence noise were introduced into the corpus, as
	described in section \ref{CNVE}.
	}
\end{sidewaysfigure*}

\end{subsection}

\end{section}

@@

\begin{section}{offcuts}
Two controlled experiments are performed:
\begin{itemize}
	\item \emph{Word frequency variation} compare the lengths of the word vectors for words with varying frequencies, but with the same co-occurrence distribution;
	\item \emph{Co-occurrence noise variation} compare the lengths of the word vectors for words with of the same frequency, but with varying levels of noise in their co-occurrence distributions.
\end{itemize}
\end{section}

\footnotesize
\bibliographystyle{te}
\bibliography{main}
\end{document}
